Comment: Reasoning about Interference Between Units

### 2014-12-02 paper/introduction.tex
Draft to Peter and Mark

@@

\subsection*{Alternative test statistics}

There are usually better ways to evaluate the fit of a model than comparing
two marginal distributions. One simple test statistic would be the
sum-of-squared-residuals (SSR) from a least squares regression of $\widetilde
y_{ij}(\0,\Z)$ on $Z_{ij}$ and $Z_{i(-j)}$. Since the model in \eqref{causal}
is linear, $\T(\widetilde y_{ij}(\0,\Z), \Z) = \sum_{ij \in U} \left[
  \widetilde \Y(\0) - (\X^T\X)^{-1}\X^T \widetilde \Y(\0) \right]^2$, where $$
\widetilde \Y(\0) \equiv \left( \begin{array}{c} \widetilde y_{11}(0) \\
    \vdots \\ y_{(N/2)(-1)}(0) \end{array} \right), \X \equiv  \left(
  \begin{array}{ccc} 1 & Z_{11} & Z_{1(-1)} \\ \vdots & \vdots & \vdots  \\ 1
    & Z_{(N/2)(-1)} & Z_{(N/2)1} \end{array} \right) = ({\bf 1} \; \Z \;
\Z^I).  $$ Assume now that either the hypothesized $\tau_1$ or $\tau_2$ is
radically incorrect: i.e., $\vert \widetilde\tau_1 - \tau_1 \vert$ or $\vert
\widetilde\tau_2 - \tau_2 \vert \gg \max(y_{ij}(\0)) - \min(y_{ij}(\0))$.
Fixing $\Z = \z$, for any randomization $\Z' \notin \{\z, (1-\z), \z^I,
(1-\z^I), \0, \one \}$, $\T(\widetilde y_{ij}(\0,\z),\Z') > \T(\widetilde
y_{ij}(\0,\z),\z)$. Assuming noncollinearity,
%$\Z \notin \{\Z^I, (1-\Z^I)\}$, , (0,...,0), (1,...,1)\}$)
$p \leq 6  \left( 2^{-N} \right)$, rejecting the hypothesis with $\alpha =
0.05$ for any $N \geq 8$. %Consistency follows straightforwardly.

In the example above, use of SSR as a test statistic has maximal power to
reject radically incorrect hypotheses, whereas the test statistics proposed by
BFP are inconsistent. But SSR from an ordinary least squares regression is not
always appropriate: for example, when the probability of exposure to spillover
is heterogeneous across individuals, we may wish to apply inverse probability
weights so as to ensure representative samples of potential outcomes. This
suggests a conjecture: that the $SSR$ from an {\it
  inverse-probability-weighted} least squares regression is more generally a
sensible test statistic for models that include interference.  (The nature of
these weights will be discussed when discussing alternative modes of
inference.) Additionally, when nonlinear deviations from model predictions are
of concern, a weighted variant of the Brownian distance covariance
\cite{szekely2009brownian} or other {\it E}-statistic may be more sensible
than SSR.

As an example of the performance of these new statistics, we re-analyze the
model and design from BFP. Recall that their model of treatment propagation
was:

@@

Briefly, the model posits that treatment effects can depend on either direct
assignment to treatment ($\bz$) governed by $\beta$ or spillover as an
increasing (but flattening) function of the number of directly connected
treated neighbors ($\bz^{T} \bS$) and is governed by $\tau$. So, we have a
model with two parameters. The network used by BFP involves 256 nodes
connected in an undirected, random graph with node degree ranging from 0 to 10
(mean degree 4, 95\% of nodes with degree between 1 and 8, five nodes with
degree 0 [i.e. unconnected]).  Treatment is assigned to 50\% of the nodes at
completely at random in the BFP example.

The top row of figure~\ref{fig:twoD} compares the power of the SSR+Design test
statistic (upper left panel) to versions of this statistic that either only
include fixed node degree (SSR+Degree) , or no information about the network
at all (SSR). For each test statistic, we tested the hypothesis
$\tau=\tau_0,\beta=\tau_0$ by using a simulated permutation test (i.e. we used
1000 permutations instead of all of them). We executed that test 10,000 times
for each pair of parameters.  The proportion of $p$-values from that test less
than .05 is plotted in Figure~\ref{fig:twoD}: darker values are fewer
rejections, lighter values record more rejections.  All of these test
statistics are valid --- they reject the true null of $\tau=.5,\beta=2$ no
more than 5\% of the time at $\alpha=.05$ --- the plots are darkest in the
area where the two lines showing the true parameters intersect. All of the
plots have some power to reject non-true alternatives --- as we can see with
the large white areas in all of the plots. However, only when we add
information about the number of treated neighbors to the SSR+Degree
statistic, do we see high power against all alternatives in the plane.

@@

The bottom row of Figure~\ref{fig:twoD} demonstrates the power of the KS test.
The bottom right hand panel shows the test used in BFP. Again all of the tests
are valid in the sense of rejecting the truth no more than 5\% of the time
when $\alpha=.05$ although all of these tests are conservative: the SSR based
tests rejected the truth roughly 4\% of the 10,000 simulations but the KS
tests rejected the truth roughly 2\% of the time.  The KS+Design and
KS+Degree panels show the power of applying the KS test to residuals from linear
models including network degree only (the +Degree version) or degree and also
the number of treated neighbors (a quantity that changes for each simulation
of the experiment). That is, where as the SSR panels used the sum of squared
residuals after accounting for network degree and/or number of treated
neighbors, the KS+Design and KS+Degree panels apply the KS test to the raw
residuals. These panels show (1) that inclusion of a quantity from the true
model (number of treated neighbors) is not enough to increase power against
all alternatives to the level shown by the SSR+Design test statistic and (2)
that the KS tests and the SSR tests have different patterns of power --- the
KS tests might be less powerful in general (more darker areas on the plots),
but our general sense from the analysis accords with the general sense about
the KS test in the scholarly literature --- that it is relatively low powered
(cites).

We cannot say here whether the SSR+Design test will provide the best power
against relevant alternatives for all possible models of treatment effect
propagation and designs. However, we hope that this research note both
improves the application of the BFP approach and raises new questions for
research.  Via counterexample, we can see that effect
increasing test statistics are not always appropriate for assessing hypotheses
about interference. BFP is correct in the assertion that, regardless of the
choice of test statistic selection, a set of implausible hypotheses is
identified by the procedure. But we should not be led to believe that, for any
given test statistic, that some hypotheses are more plausible than others.
Such inferences -- comparing hypotheses -- may depend on the test statistic
used, and not necessarily reflect the plausibility of the model at hand.

>>>>

So, say $Y_i$ is the observed outcome and we hypothesize that units do not
interfere with each other and also that $y_{i,Z_i=1}=y_{i,Z_i=0}+\tau$, we can
assess which (if any) hypothesized values of $\tau$ appear implausible from
the perspective of the data by: (1) Mapping the hypothesis about unobserved
quantities using the identity $Y_i=Z_i y_{i,Z_i=1} + (1-Z_i) y_{i,Z_i=0}$ ---
noticing that if $y_{i,Z_i=1}=y_{i,Z_i=0}+\tau$ then $y_{i,Z_i=0}=Y_i - Z_i
\tau$ (by substituting from the hypothesized relationship into the observed
data identity); (2) Using this result to adjust the observed outcome to
represent what would be implied by the hypothesis for given $\tau_0$ such that
$\widetilde y_{i,Z_i=0}=Y_i - Z_i \tau_0$; and (3) Under the hypothesis,
$\widetilde y_{i,Z_i=0}$ should have no systematic relationship with treatment
assignment, so we can summarize this relationship with a test statistic,
$\mathcal T(\widetilde y_{i,Z_i=0},Z_i)$ and the distribution of values for
this test statistic arising from repetitions of treatment assignment (new
draws of $\bz$ from all of the ways that such treatment assignment vectors
could have been produced); and finally (4) a $p$-value arises by comparing the
observed test statistic, $T(Y_i,Z_i)$ against the distribution of that test
statistic that characterizes the hypothesis.

@@

\section{Hypothesis testing as model fit assessment: The SSR test statistic}

Rosenbaum style randomization inference tends to use test statistics that
compare two distributions. Simple models imply that the distribution of the
outcome in the control remains fixed. For example, $\widetilde
y_{i,Z_i=0}=Y_i-Z_i \tau$ only changes the distribution of outcomes for units
in the treated condition. Comparing the mean of $\widetilde y_{i,Z_i=0}|Z_i=0$
to the mean of $\widetilde y_{i,Z_i=0}|Z_i=0$ makes intuitive sense in this
case, and, if $Y_i$ is Normal or at least unimodal without major outliers,
then this test might have optimal power. The complex model used as an example
by BFP involved adjustments to both control and treated outcomes --- some
hypothesized parameters would cause shifts in variance, others in location.
So, BFP proposed to use the KS-test statistic to assess the relationship
between $\widetilde y_{i,Z_i=0,Z_{-i}=0}$ and $Z_i$ (where $Z_{-i}=0$ means
"when all units other than $i$ are also not treated".

Yet, one can also think about the process of hypothesis testing as a process
of assessing model fit, and there are usually better ways to evaluate the fit
of a model than comparing two marginal distributions. In the case were we know
the fixed adjacency matrix of the network, $\bS$, and where we imagine that
network attributes (like degree) of a node play a role in the mechanism by
which treatment propagates, the idea of assessing model fit rather than
distribution comparison leads naturally to the the sum-of-squared-residuals
(SSR) from a least squares regression of $ y_{i,Z_i=0}$ on $Z_{i}$ and
$\bz^{T} \bS$ (i.e. the number of directly connected nodes assigned treatment)
as well as the $\mathbf{1}^{T} \bS$ (i.e. the degree of the node).

As an example of the performance of these new statistics, we re-analyze the
model and design from BFP. Their model of treatment propagation
was:

@@

Briefly, this model posits that treatment effects can depend on either direct
assignment to treatment ($\bz$) governed by $\beta$ or spillover as an
increasing (but flattening) function of the number of directly connected
treated neighbors ($\bz^{T} \bS$) and is governed by $\tau$. So, we have a
model with two parameters. The network used by BFP involves 256 nodes
connected in an undirected, random graph with node degree ranging from 0 to 10
(mean degree 4, 95\% of nodes with degree between 1 and 8, five nodes with
degree 0 [i.e. unconnected]).  Treatment is assigned to 50\% of the nodes at
completely at random in the BFP example.

We assess three versions of the SSR test statistic versus three versions of
the KS test statistic. The first, describe above, we call the SSR+Design test
statistic because it represents information about how treatment is assigned to
the nodes (in $\bz^T \bS$). The second version of the SSR test statistic only
includes network degree and excludes information about the treatment status of
other nodes (only includes $\b1^T \bS$). And the third version includes only
treatment assignment $\bz$. The top row of figure~\ref{fig:twoD} compares the
power of the SSR+Design test statistic (upper left panel) to versions of this
statistic that either only include fixed node degree (SSR+Degree), or no
information about the network at all (SSR). For each test statistic, we tested
the hypothesis $\tau=\tau_0,\beta=\tau_0$ by using a simulated permutation
test (i.e. we sampled 1000 permutations instead of all of them). We executed that
test 10,000 times for each pair of parameters.  The proportion of $p$-values
from that test less than .05 is plotted in Figure~\ref{fig:twoD}: darker
values show fewer rejections, lighter values record more rejections.  All of
these test statistics are valid --- they reject the true null of
$\tau=.5,\beta=2$ no more than 5\% of the time at $\alpha=.05$ --- the plots
are darkest in the area where the two lines showing the true parameters
intersect. All of the plots have some power to reject non-true alternatives
--- as we can see with the large white areas in all of the plots. However,
only when we add information about the number of treated neighbors to the
SSR+Degree statistic, do we see high power against all alternatives in the
plane.

@@

The bottom row of Figure~\ref{fig:twoD} demonstrates the power of the KS test.
The bottom right hand panel shows the test used in BFP. Again all of the tests
are valid in the sense of rejecting the truth no more than 5\% of the time
when $\alpha=.05$ although all of these tests are conservative: the SSR based
tests rejected the truth roughly 4\% of the 10,000 simulations but the KS
tests rejected the truth roughly 2\% of the time.  The KS+Design and KS+Degree
panels show the power of applying the KS test to residuals from linear models
including network degree only (the +Degree version) or degree and also the
number of treated neighbors (the +Design version). That is, whereas the SSR
panels used the sum of squared residuals after accounting for network degree
and/or number of treated neighbors, the KS+Design and KS+Degree panels apply
the KS test to the raw residuals after adjusting for information about the
design and network (or with no adjustment). These panels show (1) that
inclusion of a quantity from the true model (number of treated neighbors) is
not enough to increase power against all alternatives to the level shown by
the SSR+Design test statistic and (2) that the KS tests and the SSR tests have
different patterns of power --- the KS tests might be less powerful in general
(more darker areas on the plots), but our general sense from the analysis
accords with the general sense about the KS test in the scholarly literature
--- that it is relatively low powered.\footnote{For a short description and a
few citations to the well-known low-power of the KS-test, see
\url{https://asaip.psu.edu/Articles/beware-the-kolmogorov-smirnov-test}}

\section{Discussion and Speculations}

We cannot say here whether the SSR+Design test will provide the best power
against relevant alternatives for all possible models of treatment effect
propagation, network topologies and designs. However, we hope that this
research note both improves the application of the BFP approach and raises new
questions for research.  BFP is correct in the assertion that, regardless of
the choice of test statistic selection, a set of implausible hypotheses is
identified by the procedure. But we should not be led to believe that, for any
given test statistic, that some hypotheses are more plausible than others.
Such inferences -- comparing hypotheses -- may depend on the test statistic
used, and not necessarily reflect the plausibility of the model at hand. That
is, the results of any hypothesis test (or confidence interval creation) tell
us both about the test statistic and about the causal model under scrutiny.

In the example above, the SSR+Design test statistic had much better power than
any other test statistic. But SSR from an ordinary least squares regression is
not always appropriate: for example, when the probability of exposure to
spillover is heterogeneous across individuals and not well captured by the
$\bz^T \bS$ term or some other analogous term, we may wish to apply inverse
probability weights so as to ensure representative samples of potential
outcomes. This suggests a conjecture: that the $SSR$ from an {\it
  inverse-probability-weighted} least squares regression is more generally a
sensible test statistic for models that include
interference.\footnote{\citet{aronowsamii2012interfere} use such weights for
  unbiased estimation of network-treatment-exposure probability weighted
  average treatment effects.}  Additionally, when nonlinear deviations from
model predictions are of concern, a weighted variant of the Brownian distance
covariance \cite{szekely2009brownian} or other {\it E}-statistic may be more
sensible than SSR.
